Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-02-01 paper/main.tex
final edits through end of results

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple-choice quiz questions interleaved between
watching two course videos from the Khan Academy platform. We applied our
framework to the videos' transcripts and the text of the quiz questions to
quantify the content of each moment of video and each quiz question. We used
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings show how a
small set of quiz questions may be used to obtain rich and meaningful, high-resolution
insights into what each learner knows, and how their knowledge
changes over time as they learn.

@@

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience (e.g.,~\citealp{Kato40, Gall00, ScotEtal07, HallGree08, Macl05})
has profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
information about what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it. However, text
embedding models (e.g.,~\citealp{LandDuma97, DeerEtal90, BleiEtal03,
BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20}) also attempt to capture the
deeper meaning \textit{underlying} those atomic elements. These models consider
not only the co-occurrences of those elements within and across documents, but
also patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical properties
of the elements, and other high-level characteristics of how they are
used~\citep{Mann20, Mann21a}. According to these models, the deep conceptual
meaning of a document may be captured by a feature vector in a high-dimensional
representation space, where nearby vectors reflect conceptually related
documents. A model that succeeds at capturing an analog of ``understanding'' is
able to assign nearby feature vectors to two conceptually related documents,
\textit{even when the specific words contained in those documents have very little
overlap}.

Given these insights, what form might the representation of the sum total of a
person's knowledge take? First, we might require a means of systematically
describing or representing the nearly infinite set of possible things a person
could know. Second, we might want to account for potential associations between
different concepts. For example, the concepts of ``fish'' and ``water'' might
be associated in the sense that fish live in water. Third, knowledge may have a
critical dependency structure, such that knowing about a particular concept
might require first knowing about a set of other concepts. For example,
understanding the concept of a fish swimming in water first requires
understanding what fish and water \textit{are}. Fourth, as we learn, our
``current state of knowledge'' should change accordingly. Learning new concepts
should both update our characterizations of ``what is known'' and also
unlock any now-satisfied dependencies of those newly learned concepts so that
they are ``tagged'' as available for future learning.
% TODO: is the second half of this paragraph relevant to this paper??

Here we develop a framework for modeling how conceptual knowledge is acquired
during learning. The central idea behind our framework is to use text embedding
models to define the coordinate systems of two maps: (a) a \textit{knowledge
map} that describes the extent to which each concept is currently known and (b)
a \textit{learning map} that describes changes in knowledge over time. Each
location on these maps represents a single concept, and the maps' geometries
are defined such that related concepts are located nearby in space. We use this
framework to analyze and interpret behavioral data collected from an experiment
that had participants answer sets multiple-choice questions about a series of
recorded course lectures.

@@

At its core, our main modeling approach is based around a simple assumption
that we sought to test empirically: all else being equal, knowledge about a
given concept is predictive of knowledge about similar or related concepts.
From a geometric perspective, this assumption implies that knowledge is
fundamentally ``smooth.'' In other words, as one moves through a space
representing an individual's knowledge (where similar concepts occupy nearby
coordinates), their ``level of knowledge'' should change relatively gradually
throughout that space. To begin to test this smoothness assumption, we sought
to track participants' knowledge and how it changed over time in response to
training. Two overarching goals guide our approach. First, we want to gain
detailed insights into what learners know, at different points in their
training. For example, rather than simply reporting on the proportions of
questions participants answer correctly (i.e., their overall performance), we
seek estimates of their knowledge about a variety of specific concepts.
Second, we want our approach to be potentially scalable to large numbers of
concepts, courses, and students. This requires the conceptual content of
interest to be discovered \textit{automatically}, rather than relying on
manually produced ratings or labels.

@@

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics not specifically presented in either
video (see Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(lecture 1, lecture 2, and general physics knowledge) on each of the three
quizzes. Quiz 1 was intended to assess participants' ``baseline'' knowledge
before training, quiz 2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., lecture 1), and quiz 3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., lecture 2).

@@

\caption{\textbf{Modeling course content.} \textbf{A. Building a document pool
from sliding windows of text.} We decompose each lecture's transcript into a
series of overlapping sliding windows. The set of transcript snippets (across
all windows) may be treated as a set of ``documents'' for training a text
embedding model. \textbf{B. Constructing lecture content
\textit{trajectories}.} After training our model on the sliding windows from
both lectures, we transform each lecture into a ``trajectory'' through text
embedding space by joining the embedding coordinates of successive sliding
windows parsed from its transcript. \textbf{C. Embedding multiple lectures and
questions in a shared space.} We apply the same model (trained on the two
lectures' windows) to both lectures, along with the text of each question in
our pool (Tab.~\questions), to project them into a shared text embedding space.
This results in one trajectory per lecture and one coordinate for each
question. Here we have projected the 15-dimensional embeddings onto their first
3 principal components for visualization.}

@@

To study in detail how participants' conceptual knowledge changed over the
course of the experiment, we first sought to model the conceptual content
presented to them at each moment throughout each of the two lectures. We
adapted an approach we developed in prior work~\citep{HeusEtal21} to identify
the latent themes in the lectures using a topic model~\citep{BleiEtal03}.
Briefly, topic models take as input a collection of text documents and learn a
set of ``topics'' (i.e., latent themes) from their contents. Once fit, a topic
model can be used to transform arbitrary (potentially new) documents into sets
of ``topic proportions,'' describing the weighted blend of learned topics
reflected in their texts. We parsed automatically generated transcripts of the
two lectures into overlapping sliding windows, where each window contained the
text of the lecture transcript from a particular time range. We treated the set
of text snippets (across all of these windows) as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming the text from every sliding window with our model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix that
described the unique mixture of broad themes from both lectures reflected in
each window's text. Each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is a coordinate in a 15-dimensional space whose axes
are topics discovered by the model. Within this space, each lecture's sequence
of topic vectors (i.e., corresponding to its transcript's overlapping text
snippets across sliding windows) forms a \textit{trajectory} that captures how
its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We
resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices) then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question sets (averaging across time and questions, respectively) reveals a
striking correspondence (Fig.~\topicWeights). Specifically, the average topic
weights from lecture 1 are strongly correlated with the average topic weights
from lecture 1 questions ($r(13) = 0.809, p < 0.001$, 95\% confidence interval
(CI)~$= [0.633, 0.962]$), and the average topic weights from lecture 2 are
strongly correlated with the average topic weights from lecture 2 questions
($r(13) = 0.728, p = 0.002, CI = [0.456, 0.920]$). At the same time, the
average topics from two lectures are \textit{negatively} correlated ($r(13) =
-0.634, p = 0.011, CI = [-0.924, -0.237]$), indicating that the topic model
also exhibits some degree of specificity. The full set of pairwise comparisons
between topic vectors for the lectures and each question set is reported in
Figure~\topicWeights.

@@

\caption{\textbf{Lecture and question topic overlap.} \textbf{A. Topic weight
variability}. The bar plots display the variance of each topic's weight across
lecture timepoints (top row) and questions (bottom row); colors denote topics.
The top-weighted words from the most ``expressive'' (i.e., variable across
observations) topic from each lecture are displayed in the upper right (orange:
topic 2; yellow-green: topic 4). The top-weighted words from the full set of
topics may be found in Table~\topics. \textbf{B. Relationships between topic
weight variability.} Pairwise correlations between the distributions of topic
weight variance for each lecture and question category. Each row and column
corresponds to a bar plot in Panel A.}

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or questions) reflect about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variances in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the lecture 1 video
and questions ($r(13) = 0.824,~p<0.001, CI = [0.696, 0.973]$) and the lecture 2
video and questions ($r(13) = 0.801,~p<0.001,~95\%~\mathrm{CI} = [0.539,
0.958]$). However, as reported in Figure~\ref{fig:topics}B, the variability in
topic expressions across \textit{different} videos and lecture-specific
questions (i.e., lecture 1 video versus lecture 2 questions; lecture 2 video
versus lecture 1 questions) were negatively correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figures~\ref{fig:topics} and~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondances (at a
coarse scale) between the lectures and (held-out) questions.

Although a single lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether the text embeddings could be used to
automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors from each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also examined the
best-matching intervals for each question qualitatively by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question covered and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

@@

The ability to quantify how much each question is ``asking about'' the content
from each moment of the lectures could enable high-resolution insights into
participants' knowledge. Traditional approaches to estimating how much a
student ``knows'' about the content of a given lecture entail computing the
proportion of correctly answered questions. But if two students receive
identical scores on an exam, might our modeling framework help us to gain more
nuanced insights into the \textit{specific} content that each student has
mastered (or failed to master)? For example, a student who misses three
questions that were all about the same concept (e.g., concept $A$) will have
gotten the same \textit{proportion} of questions correct as another student who
missed three questions about three \textit{different} concepts (e.g., $A$, $B$,
and $C$). But if we wanted to fill in the ``gaps'' in the two students'
understandings, we might do well to focus on concept $A$ for the first student,
but to also add in materials pertaining to concepts $B$ and $C$ for the second
student. In other words, raw ``proportion-correct'' measures may capture
\textit{how much} a student knows, but not \textit{what} they know. We wondered
whether our modeling framework might enable us to (formally and automatically)
infer participants' knowledge at the scale of individual concepts (e.g., as
captured by a single question).

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \nameref{subsec:traces}). Essentially, the
estimated knowledge at the coordinate is given by the weighted average
proportion of quiz questions the participant answered correctly, where the
weights reflect how much each question is ``about'' the content at $x$. When we
apply this approach to estimate the participant's knowledge about the content
presented in each moment of each lecture, we can obtain a detailed timecourse
describing how much ``knowledge'' the participant has about any part of the
lecture. As shown in Figure~\ref{fig:knowledge-timeseries}, we can also apply
this approach separately for the questions from each quiz the participants took
throughout the experiment. From just 13 questions per quiz, we obtain a
high-resolution snapshot (at the time each quiz was taken) of what the
participants knew about any moment's content, from either of the two lectures
they watched (comprising a total of 1106 samples across the two lectures).

@@

Of course, even though the timecourses in
Figure~\ref{fig:knowledge-timeseries}A and C provide detailed
\textit{estimates} about participants' knowlege, those estimates are only
\textit{useful} to the extent that they accurately reflect what participants
actually know. As one sanity check, we anticipated that the knowledge estimates
should show a content-specific ``boost'' in participants' knowledge after
watching each lecture. In other words, if participants learn about each
lecture's content when they watch each lecture, the knowledge estimates should
reflect that. After watching the \textit{Four Fundamental Forces} lecture,
participants should show more knowledge for the content of that lecture than
they had before, and that knowledge should persist for the remainder of the
experiment. Specifically, knowledge about that lecture's content should be
relatively low when estimated using Quiz 1 responses, but should increase when
estimated using Quiz 2 or 3 responses (Fig.~\ref{fig:knowledge-timeseries}B).
Indeed, we found that participants' estimated knowledge about the content of
the \textit{Four Fundamental Forces} was substantially higher on Quiz 2 versus
Quiz 1 ($t(49) = 8.764, p < 0.001$) and on Quiz 3 versus Quiz 1 ($t(49) =
10.519, p < 0.001$). We found no reliable differences in estimated knowledge
about that lecture's content on Quiz 2 versus 3 ($t(49) = 0.160, p = 0.874$).
Similarly, we hypothesized (and subsequently confirmed) that participants
should show more estimated knowledge about the content of the \textit{Birth of
Stars} lecture after (versus before) watching it
(Fig.~\ref{fig:knowledge-timeseries}D). Specifically, since participants
watched that lecture after taking Quiz 2 (but before Quiz 3), we hypothesized
that their knowledge estimates should be relatively low on Quizzes 1 and 2, but
should show a ``boost'' on Quiz 3. Consistent with this prediction, we found no
reliable differences in estimated knowledge about the \textit{Birth of Stars}
lecture content on Quizzes 1 versus 2 ($t(49) = 1.013, p = 0.316$), but the
estimated knowledge was substantially higher on Quiz 3 versus 2 ($t(49) =
10.561, p < 0.001$) and Quiz 3 versus 1 ($t(49) = 8.969, p < 0.001$).

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, the estimated knowledge should have some
predictive information about whether the participant is likely to answer the
question correctly or incorrectly. For each question in turn, for each
participant, we used Equation~\ref{eqn:prop} to estimate (using all
\textit{other} questions from the same quiz, from the same participant) the
participant's knowledge at the held-out question's embedding coordinate. For
each quiz, we grouped these estimates into two distributions: one for the
estimated knowledge at the coordinates of each \textit{correctly} answered
question, and another for the estimated knowledge at the coordinates of each
\textit{incorrectly} answered question (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

    \caption{\textbf{Estimating knowledge at the embedding coordinates of
    held-out questions.} Separately for each quiz (panel), we plot the
    distributions of predicted knowledge at the embedding coordinates of each
    held out correctly (blue) or incorrectly (red) answered question. The
    $t$-tests reported in each panel are between the distributions of estimated
    knowledge at the coordinates of correctly versus incorrectly answered
    held-out questions.}

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) into a shared 2-dimensional space (see
\nameref{subsec:knowledge-maps}). Next, we sampled points from a $100 \times
100$ grid of coordinates that evenly tiled a rectangle enclosing the 2D
projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10,000 sampled locations, and we averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted text embedding
model.

>>>>

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high-dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple-choice quiz questions, interleaved between
watching two course videos from the Khan Academy platform. We apply our
framework to the videos' transcripts and the text of the quiz questions to
quantify the content of each moment of video and each quiz question. We use
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings show how a
small set of quiz questions may be used to obtain rich and meaningful, high-resolution
insights into what each learner knows, and how their knowledge
changes over time as they learn.

@@

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience (e.g.,~\citealp{Kato40, Gall00, ScotEtal07, HallGree08, Macl05})
has profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
clues as to what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it. However, text
embedding models (e.g.,~\citealp{LandDuma97, DeerEtal90, BleiEtal03,
BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20}) also attempt to capture the
deeper meaning \textit{underlying} those atomic elements. These models consider
not only the co-occurrences of those elements within and across documents, but
also patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical properties
of the elements, and other high-level characteristics of how they are
used~\citep{Mann20, Mann21a}. According to these models, the deep conceptual
meaning of a document may be captured by a feature vector in a high-dimensional
representation space, wherein nearby vectors reflect conceptually related
documents. A model that succeeds at capturing an analogue of ``understanding'' is
able to assign nearby feature vectors to two conceptually related documents,
\textit{even when the specific words contained in those documents have very little
overlap}.

Given these insights, what form might a representation of the sum total of a
person's knowledge take? First, we might require a means of systematically
describing or representing the nearly infinite set of possible things a person
could know. Second, we might want to account for potential associations between
different concepts. For example, the concepts of ``fish'' and ``water'' might
be associated in the sense that fish live in water. Third, knowledge may have a
critical dependency structure, such that knowing about a particular concept
might require first knowing about a set of other concepts. For example,
understanding the concept of a fish swimming in water first requires
understanding what fish and water \textit{are}. Fourth, as we learn, our
``current state of knowledge'' should change accordingly. Learning new concepts
should both update our characterizations of ``what is known'' and also
unlock any now-satisfied dependencies of those newly learned concepts so that
they are ``tagged'' as available for future learning.
% TODO: is the second half of this paragraph relevant to this paper??

Here we develop a framework for modeling how conceptual knowledge is acquired
during learning. The central idea behind our framework is to use text embedding
models to define the coordinate systems of two maps: a \textit{knowledge
map} that describes the extent to which each concept is currently known, and
a \textit{learning map} that describes changes in knowledge over time. Each
location on these maps represents a single concept, and the maps' geometries
are defined such that related concepts are located nearby in space. We use this
framework to analyze and interpret behavioral data collected from an experiment
that had participants answer sets multiple-choice questions about a series of
recorded course lectures.

@@

At its core, our main modeling approach is based around a simple assumption
that we sought to test empirically: all else being equal, knowledge about a
given concept is predictive of knowledge about similar or related concepts.
From a geometric perspective, this assumption implies that knowledge is
fundamentally ``smooth.'' In other words, as one moves through a space
representing an individual's knowledge (where similar concepts occupy nearby
coordinates), their ``level of knowledge'' should change relatively gradually
throughout that space. To begin to test this smoothness assumption, we sought
to track participants' knowledge and how it changed over time in response to
training. Two overarching goals guide our approach. First, we want to gain
detailed insights into what learners know, at different points in their
training. For example, rather than simply reporting on the proportions of
questions participants answer correctly (i.e., their overall performance), we
seek estimates of their knowledge about a variety of specific concepts.
Second, we want our approach to be potentially scalable to large numbers of
concepts, courses, and students. This requires that the conceptual content of
interest be discovered \textit{automatically}, rather than relying on
manually produced ratings or labels.

@@

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics not specifically presented in either
video (see Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(lecture 1, lecture 2, and general physics knowledge) on each of the three
quizzes. Quiz 1 was intended to assess participants' ``baseline'' knowledge
before training, Quiz 2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., lecture 1), and Quiz 3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., lecture 2).

@@

\caption{\textbf{Modeling course content.} \textbf{A. Building a document pool
from sliding windows of text.} We decompose each lecture's transcript into a
series of overlapping sliding windows. The full set of transcript snippets (across
all windows) may be treated as a set of ``documents'' for training a text
embedding model. \textbf{B. Constructing lecture content
\textit{trajectories}.} After training our model on the sliding windows from
both lectures, we transform each lecture into a ``trajectory'' through text
embedding space by joining the embedding coordinates of successive sliding
windows parsed from its transcript. \textbf{C. Embedding multiple lectures and
questions in a shared space.} We apply the same model (trained on the two
lectures' windows) to both lectures, along with the text of each question in
our pool (Tab.~\questions), to project them into a shared text embedding space.
This results in one trajectory per lecture and one coordinate for each
question. Here, we have projected the 15-dimensional embeddings onto their first
3 principal components for visualization.}

@@

To study in detail how participants' conceptual knowledge changed over the
course of the experiment, we first sought to model the conceptual content
presented to them at each moment throughout each of the two lectures. We
adapted an approach we developed in prior work~\citep{HeusEtal21} to identify
the latent themes in the lectures using a topic model~\citep{BleiEtal03}.
Briefly, topic models take as input a collection of text documents, and learn a
set of ``topics'' (i.e., latent themes) from their contents. Once fit, a topic
model can be used to transform arbitrary (potentially new) documents into sets
of ``topic proportions,'' describing the weighted blend of learned topics
reflected in their texts. We parsed automatically generated transcripts of the
two lectures into overlapping sliding windows, where each window contained the
text of the lecture transcript from a particular time span. We treated the set
of text snippets (across all of these windows) as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming the text from every sliding window with our model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix that
described the unique mixture of broad themes from both lectures reflected in
each window's text. Each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional space whose axes
are topics discovered by the model. Within this space, each lecture's sequence
of topic vectors (i.e., corresponding to its transcript's overlapping text
snippets across sliding windows) forms a \textit{trajectory} that captures how
its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We
resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices), then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question set (averaging across time and questions, respectively) reveals a
striking correspondence (Fig.~\topicWeights). Specifically, the average topic
weights from lecture 1 are strongly correlated with the average topic weights
from lecture 1 questions ($r(13) = 0.809,~p < 0.001$, 95\% confidence interval
(CI)~$= [0.633,~0.962]$), and the average topic weights from lecture 2 are
strongly correlated with the average topic weights from lecture 2 questions
($r(13) = 0.728,~p = 0.002$, 95\% CI~$= [0.456,~0.920]$). At the same time, the
average topic weights from the two lectures are \textit{negatively} correlated with 
their non-matching question sets (lecture 1 video vs.~lecture 2 questions: $r(13) = 
-0.547,~p = 0.035$, 95\% CI~$= [-0.812, ~-0.231]$; lecture 2 video vs.~lecture 1 questions: $r(13) 
= -0.612,~p = 0.015$, 95\% CI~$= [-0.874,~-0.281]$), indicating that the topic model
also exhibits some degree of specificity. The full set of pairwise comparisons
between average topic weights for the lectures and question sets is reported in
Figure~\topicWeights.

@@

\caption{\textbf{Lecture and question topic overlap.} \textbf{A. Topic weight
variability}. The bar plots display the variance of each topic's weight across
lecture timepoints (top row) and questions (bottom row); colors denote topics.
The top-weighted words from the most ``expressive'' (i.e., variable across
observations) topic from each lecture are displayed in the upper right (orange:
topic 2; yellow-green: topic 4). The top-weighted words from the full set of
topics may be found in Table~\topics. \textbf{B. Relationships between topic
weight variability.} Pairwise correlations between the distributions of topic
weight variance for each lecture and question set. Each row and column
corresponds to a bar plot in Panel A.}

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or question set) reflects about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variances in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the lecture 1 video
and questions ($r(13) = 0.824,~p<0.001$, 95\% CI~$= [0.696,~0.973]$) and the lecture 2
video and questions ($r(13) = 0.801,~p<0.001$, 95\% CI~$= [0.539,~0.958]$). 
However, as reported in Figure~\ref{fig:topics}B, the variability in
topic expressions across \textit{different} videos and lecture-specific
questions (i.e., lecture 1 video vs.~lecture 2 questions; lecture 2 video
vs.~lecture 1 questions) were negatively correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figures~\ref{fig:topics} and~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondences (at a
coarse scale) between the lectures and questions.

Although a single lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether the text embeddings could be used to
automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors from each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also qualitatively examined the
best-matching intervals for each question by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question probed and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

@@

The ability to quantify how much each question is ``asking about'' the content
from each moment of the lectures could enable high-resolution insights into
participants' knowledge. Traditional approaches to estimating how much a
student ``knows'' about the content of a given lecture entail computing the
proportion of correctly answered questions. But if two students receive
identical scores on an exam, might our modeling framework help us to gain more
nuanced insights into the \textit{specific} content that each student has
mastered (or failed to master)? For example, a student who misses three
questions that were all about the same concept (e.g., concept $A$) will have
gotten the same \textit{proportion} of questions correct as another student who
missed three questions about three \textit{different} concepts (e.g., $A$, $B$,
and $C$). But if we wanted to fill in the ``gaps'' in the two students'
understandings, we might do well to focus on concept $A$ for the first student,
but to also add in materials pertaining to concepts $B$ and $C$ for the second
student. In other words, raw ``proportion-correct'' measures may capture
\textit{how much} a student knows, but not \textit{what} they know. We wondered
whether our modeling framework might enable us to (formally and automatically)
infer participants' knowledge at the scale of individual concepts (e.g., as
captured by a single moment of a lecture).

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \nameref{subsec:traces}). Essentially, the
estimated knowledge at the coordinate is given by the weighted average
proportion of quiz questions the participant answered correctly, where the
weights reflect how much each question is ``about'' the content at $x$. When we
apply this approach to estimate the participant's knowledge about the content
presented in each moment of each lecture, we can obtain a detailed timecourse
describing how much ``knowledge'' the participant has about any part of the
lecture. As shown in Figure~\ref{fig:knowledge-timeseries}, we can also apply
this approach separately for the questions from each quiz the participants took
throughout the experiment. From just a few questions per quiz, we obtain a
high-resolution snapshot (at the time each quiz was taken) of what the
participants knew about any moment's content, from either of the two lectures
they watched (comprising a total of 1,100 samples across the two lectures).

@@

Of course, even though the timecourses in
Figure~\ref{fig:knowledge-timeseries}A and C provide detailed
\textit{estimates} about participants' knowlege, those estimates are only
\textit{useful} to the extent that they accurately reflect what participants
actually know. As one sanity check, we anticipated that the knowledge estimates
should show a content-specific ``boost'' in participants' knowledge after
watching each lecture. In other words, if participants learn about each
lecture's content when they watch each lecture, the knowledge estimates should
reflect that. After watching the \textit{Four Fundamental Forces} lecture,
participants should show more knowledge for the content of that lecture than
they had before, and that knowledge should persist for the remainder of the
experiment. Specifically, knowledge about that lecture's content should be
relatively low when estimated using Quiz 1 responses, but should increase when
estimated using Quiz 2 or 3 responses (Fig.~\ref{fig:knowledge-timeseries}B).
Indeed, we found that participants' estimated knowledge about the content of
the \textit{Four Fundamental Forces} was substantially higher on Quiz 2 versus
Quiz 1 ($t(49) = 8.764,~p < 0.001$) and on Quiz 3 versus Quiz 1 ($t(49) =
10.519,~p < 0.001$). We found no reliable differences in estimated knowledge
about that lecture's content on Quiz 2 versus 3 ($t(49) = 0.160,~p = 0.874$).
Similarly, we hypothesized (and subsequently confirmed) that participants
should show more estimated knowledge about the content of the \textit{Birth of
Stars} lecture after (versus before) watching it
(Fig.~\ref{fig:knowledge-timeseries}D). Specifically, since participants
watched that lecture after taking Quiz 2 (but before Quiz 3), we hypothesized
that their knowledge estimates should be relatively low on Quizzes 1 and 2, but
should show a ``boost'' on Quiz 3. Consistent with this prediction, we found no
reliable differences in estimated knowledge about the \textit{Birth of Stars}
lecture content on Quizzes 1 versus 2 ($t(49) = 1.013,~p = 0.316$), but the
estimated knowledge was substantially higher on Quiz 3 versus 2 ($t(49) =
10.561,~p < 0.001$) and Quiz 3 versus 1 ($t(49) = 8.969,~p < 0.001$).

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should carry some
predictive information about whether the participant is likely to answer the
question correctly or incorrectly. We developed a statistical approach to test this claim. 
For each question in turn, for each
participant, we used Equation~\ref{eqn:prop} to estimate (using all
\textit{other} questions from the same quiz, from the same participant) the
participant's knowledge at the held-out question's embedding coordinate. For
each quiz, we grouped these estimates into two distributions: one for the
estimated knowledge at the coordinates of each \textit{correctly} answered
question, and another for the estimated knowledge at the coordinates of each
\textit{incorrectly} answered question (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

    \caption{\textbf{Estimating knowledge at the embedding coordinates of
    held-out questions.} Separately for each quiz (panel), we plot the
    distributions of predicted knowledge at the embedding coordinates of each
    held-out correctly (blue) or incorrectly (red) answered question. The
    $t$-tests reported in each panel are between the distributions of estimated
    knowledge at the coordinates of correctly versus incorrectly answered
    held-out questions.}

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) onto a shared 2-dimensional plane (see
\nameref{subsec:knowledge-maps}). Next, we sampled points from a $100 \times
100$ grid of coordinates that evenly tiled a rectangle enclosing the 2D
projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10,000 sampled locations, and averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}A). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted text embedding
model.
